---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/blocklist)](https://CRAN.R-project.org/package=blocklist)
[![R-CMD-check](https://github.com/hypertidy/blocklist/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/hypertidy/blocklist/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->
  
  
# blocklist

`blocklist` builds virtual **reference stores** from collections of
NetCDF4/HDF5 files - references record where a block (or chunk) of array data
starts and ends within a file. It composes the files with GDAL's
`gdal mdim mosaic` to get the logical layout, then reads each chunk's byte
offset with rhdf5 to get the physical layout, and writes a kerchunk-Parquet
store. The result is readable by xarray, the GDAL Zarr driver, and is intended
as an intermiediate input to an Icechunk publish step. Intermediate steps
is a primary motivation of this project, enabling broader software language support
and options for use. 

## How it works

- **GDAL composes the logical array.** `gdal mdim mosaic` stitches many files
  into one VRT: dimension order (C order), dtype, CF `scale`/`offset`/`nodata`,
  composed coordinate values, and a `source -> DestSlab` map saying which file
  supplies which slab. No data is read when generating the VRT. 
- **rhdf5 supplies the physical layer.** For each source array, `h5getAllChunkInfo`
  walks the HDF5 chunk index and returns each chunk's byte address and length
  (metadata only - chunk data is never read). The codec and storage chunk shape
  come from the creation property list (`H5Pget_filter`, `H5Pget_chunk`).
- **The store is kerchunk-Parquet.** A `.zmetadata` plus per-variable
  `refs.<N>.parq` shards - fsspec's `LazyReferenceMapper` layout, Zarr v2. This
  *is* the portable sibling and the input to VirtualiZarr.

## What the heck?

Motivation: VirtualiZarr fuses three things into one in-memory object: the lazy concatenation logic, the virtual reference store, and the file addressing. We want these things to be clean and separable and not bundled into one-way-is-the-right-way mantra. 

The **concat/placement** logic is the VRT. A GDAL cli or Run job expressed as logical slab offsets in XML, not a Python object graph. It's inspectable, save-able, and we can open as-xarray with gdalxarray 
The **byte refs** are a *flat table* - (coord, addr, size, path) rows - so compute per-source, in parallel, with no shared state, and rbind. No entanglement in a lazy python array of somethings. 
As rows, the mirai version is trivial: map -> rows -> rbind, no serializing live objects.
The paths are two columns, `access`/`public`  old-school "where's the file" and we can re-point with byte refs to either however we like. The table is queryable, the array indexing logic is implicit but robust (we need explicit refs and separation for true query pushdown, but that's relatively simple). 

It's a *table*, we can store anything with the refs - that could be a vector-driver-based meta-array-driver (like GTI), or a baked-in query scheme in whatever interface you like. 

## Sources: `access` and `public`

A source file has two roles. `access` is the path that gets *opened* - handed to
`gdal mdim mosaic` and opened by rhdf5 to scan - and is the join key back to the
VRT. `public` is the durable URL *written into the references*; nothing opens it
at build time. Because chunk byte offsets are identical in any byte-identical
copy of a file, the two may differ: scan a cheap local/NFS path, reference a
remote URL.

```{r, eval=FALSE}
library(blocklist)

urls <- sprintf(
  "https://www.ncei.noaa.gov/data/sea-surface-temperature-optimum-interpolation/v2.1/access/avhrr/198109/oisst-avhrr-v02r01.1981%02d%02d.nc",
  9, 1:10)
nfs  <- sub("^https://www.ncei.noaa.gov/data/",
            "/rdsi/PUBLIC/raad/data/www.ncei.noaa.gov/data/", urls)

src <- mosaic_sources(public = urls, access = nfs)   # access = public for all-remote
```

## Example (OISST AVHRR daily)

Note we're inlining file paths to a '%s' bash call, we probably want writeLines to file and '@filelist' per `gdal mdim mosaic @filelist.txt`. 

```{r,eval=FALSE}
# 1. GDAL composes the logical manifest from the access paths
system(sprintf("gdal mdim mosaic %s mosaic.vrt", paste(src$access, collapse = " ")))

# 2. blocklist reads byte offsets and writes the reference store
virtualize_mosaic("mosaic.vrt", "oisst.zarr", sources = src)

# 3. GDAL reads it back as a Zarr datacube
system("gdal mdim info ZARR:oisst.zarr")
```

```{r,eval=FALSE}
# the references: durable URLs + per-chunk byte address/length
arrow::read_parquet("oisst.zarr/anom/refs.0.parq")
#> # A tibble: n x 4
#>   path                              offset  size   raw
#>   <chr>                              <dbl> <dbl> <blob>
#> 1 https://.../oisst-...19810901.nc   47747 684063  NA
#> 2 https://.../oisst-...19810902.nc   47747 686733  NA
```

Then in Python, the same store feeds VirtualiZarr, xarray, and Icechunk:

```python
from virtualizarr import open_virtual_dataset
from virtualizarr.parsers import KerchunkParquetParser

vds = open_virtual_dataset("oisst.zarr", parser=KerchunkParquetParser())
# vds.vz.to_icechunk(store, config=ManifestSplittingConfig(...))   # split along time
```

## Notes 

- **Offsets are copy-invariant.** A chunk's byte address is intrinsic to the
  HDF5 layout, so you can scan any copy and still reference the canonical URL.
- **C order, everywhere.** `H5Pget_chunk` and `h5getAllChunkInfo` return C order;
  `h5ls()$dim` is *reversed* for R display (use `native = TRUE` to keep C order).
  Mixing the two scrambles the chunk-index maths silently.
- **`gzip`, not `zlib`.** The Zarr-spec compressor id for DEFLATE is `gzip`. GDAL
  accepts only spec ids; zarr-python/numcodecs also accept `zlib`. Emit `gzip`
  and both readers work. Decompression ignores the level, so it's irrelevant.
- **Codec homogeneity is a hard boundary.** A non-zero HDF5 filter mask on any
  chunk, or a changed `scale`/`offset` across files, means the span can't be one
  Zarr array - the same constraint that splits long satellite archives into eras.
- **The VRT drops CF units.** `gdal mdim mosaic` carries coordinate *values* but
  not all attributes (e.g. `time` loses `units`); read those from a source file.
- **Open read-only.** `h5default("H5F_ACC_RD")` is `"H5F_ACC_RDWR"` - pass
  `flags = "H5F_ACC_RDONLY"` explicitly, or opens fail on read-only mounts.
- **parquet partitions** these files are numbered in ordinal manner, not sorted.

## What the store is, in Parquet terms

A directory of independent Parquet files, manually **sharded** by variable
(subdirectory) and by flat-chunk-index range (`refs.<N>.parq`, `record_size`
rows each), with the partition keys implied **positionally** - `<var>` from the
path, chunk index from `N * record_size + row` - not stored as columns or
Hive-encoded. It is *not* an Arrow/Hive partitioned dataset; the index lives in
the filesystem layout, the way Zarr keys do. Sharding bounds per-file size and
enables lazy per-shard reads - the read-side dual of Icechunk manifest splitting.

This is what 16344 netcdf files of OISST looks like on disk: 

```bash
du -d 1 -h oisst_test.zarr/
8.0K    oisst_test.zarr/zlev
264K    oisst_test.zarr/anom
328K    oisst_test.zarr/sst
12K     oisst_test.zarr/lon
336K    oisst_test.zarr/ice
12K     oisst_test.zarr/lat
332K    oisst_test.zarr/err
68K     oisst_test.zarr/time
1.4M    oisst_test.zarr/


Rscript -e "tibble::as_tibble(arrow::read_parquet('oisst_test.zarr/sst/refs.0.parq'))"
# A tibble: 100,000 × 4
   path                                                      offset   size   raw
   <chr>                                                      <int>  <int> <arr>
 1 https://www.ncei.noaa.gov/data/sea-surface-temperature-… 1025881 683590
 2 https://www.ncei.noaa.gov/data/sea-surface-temperature-… 1031813 684863
 3 https://www.ncei.noaa.gov/data/sea-surface-temperature-… 1032012 683768
 4 https://www.ncei.noaa.gov/data/sea-surface-temperature-… 1035853 682292
 5 https://www.ncei.noaa.gov/data/sea-surface-temperature-… 1032925 681529
 6 https://www.ncei.noaa.gov/data/sea-surface-temperature-… 1029001 681074
 7 https://www.ncei.noaa.gov/data/sea-surface-temperature-… 1027366 680356
 8 https://www.ncei.noaa.gov/data/sea-surface-temperature-… 1028374 680184
 9 https://www.ncei.noaa.gov/data/sea-surface-temperature-… 1026568 679206
10 https://www.ncei.noaa.gov/data/sea-surface-temperature-… 1029553 678224
# ℹ 99,990 more rows

```

## what this could be

A `gdal mdim index` application, there's a pending draft PR but we also need the
fast chunk iteration from HDF5 itself ported into GDAL. No aspirations for other formats. 


## Code of Conduct

Please note that the blocklist project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/1/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
